---
title: "Generative validation of SSF / iSSF models"
output: rmarkdown::html_vignette
embed-resources: true
vignette: >
  %\VignetteIndexEntry{Generative validation of SSF / iSSF models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4,
  message = FALSE,
  warning = FALSE
)

run_deer_analysis <- identical(
  tolower(Sys.getenv("GMOV_RUN_DEER_VIGNETTE")),
  "true"
) &&
  requireNamespace("amt", quietly = TRUE) &&
  requireNamespace("terra", quietly = TRUE) &&
  requireNamespace("survival", quietly = TRUE) &&
  requireNamespace("mvtnorm", quietly = TRUE)

deer_n_sims <- suppressWarnings(as.integer(Sys.getenv("GMOV_DEER_N_SIMS", "9")))
if (!is.finite(deer_n_sims) || deer_n_sims < 2L) {
  deer_n_sims <- 9L
}
```

```{r setup}
library(gmov)
```

## Purpose

Step-selection functions (SSFs) and integrated step-selection analyses (iSSAs)
are usually fitted to one-step movement decisions. A fitted model can therefore
describe local choices well while still generating unrealistic trajectories when
it is iterated forward. The goal of `gmov` is to validate supplied simulated
trajectories against the observed trajectory using interpretable, trajectory
level diagnostics.

The framework follows Nicosia (2026), which evaluates generative realism with
four complementary pillars:

1. Emergent space use: compare observed and simulated utilization distributions.
2. Diffusion behavior: compare mean squared displacement curves.
3. Path structure: compare path straightness or sinuosity.
4. Barrier interactions: compare intersections with a known linear barrier.

`gmov` implements these diagnostics after simulation. It does not fit SSF or
iSSF models and it does not replace `amt`. A typical workflow is:

1. prepare tracks and steps with `amt`;
2. fit an SSF or iSSF model with `amt`;
3. simulate replicate trajectories from the fitted model;
4. pass the observed and simulated tracks to `gmov`.

The Monte Carlo p-values reported by `gmov` are conditional diagnostics for the
fitted model and simulation procedure supplied by the user. They should be read
as checks of generative compatibility, not as universal goodness-of-fit
guarantees.

## Red deer case study from the article

The empirical application in Nicosia (2026) uses the red deer trajectory and
forest raster distributed with `amt`. The data describe one red deer (*Cervus
elaphus*) from northern Germany. The article uses a regular 6-hour trajectory,
a binary forest covariate, a 1:25 used-to-available step design, and a standard
iSSA with forest selection and movement-kernel terms.

The publication-scale analysis uses 99 simulated trajectories, giving a minimum
Monte Carlo p-value of 0.01 under the +1 correction. The article reports partial
generative realism for this individual: the fitted iSSA reproduces broad
space-use and path straightness, but the mean squared displacement diagnostic
detects a dynamic mismatch. Because no explicit barrier geometry is specified
for this empirical case study, only the first three pillars are used.

The code below is the package vignette version of that workflow. It is complete,
but it is not run automatically during package checks because it depends on a
working `amt` installation and can be computationally expensive. To run it,
render the vignette with:

```r
Sys.setenv(GMOV_RUN_DEER_VIGNETTE = "true", GMOV_DEER_N_SIMS = "99")
devtools::build_vignettes()
```

```{r deer-skip-note, echo = FALSE, results = "asis"}
if (!run_deer_analysis) {
  cat(
    "The executable red deer analysis chunks are skipped in this build. ",
    "Set `GMOV_RUN_DEER_VIGNETTE=true` and use an R environment where `amt`, ",
    "`terra`, `survival`, and `mvtnorm` load correctly to run the full workflow.\n",
    sep = ""
  )
}
```

## Load the red deer data

The `amt::deer` object is already a track with `x_`, `y_`, and `t_` columns in
EPSG:3035. The accompanying `amt::sh_forest` raster stores binary forest cover
for the study area.

```{r deer-data, eval = run_deer_analysis}
library(amt)
library(terra)
library(survival)

set.seed(2025)

data("deer", package = "amt")
data("sh_forest", package = "amt")

forest_map <- terra::rast(sh_forest)
names(forest_map) <- "forest"

data.frame(
  n_fixes = nrow(deer),
  start = min(deer$t_),
  end = max(deer$t_),
  median_interval_hours = stats::median(as.numeric(diff(deer$t_), units = "hours"))
)
```

The article resamples the track to a regular 6-hour interval with 30 minutes of
tolerance and keeps bursts with at least two locations.

```{r prepare-track, eval = run_deer_analysis}
observed_track <- deer |>
  amt::track_resample(rate = amt::hours(6), tolerance = amt::minutes(30)) |>
  amt::filter_min_n_burst(min_n = 2)

observed_track <- observed_track[, c("x_", "y_", "t_")]

data.frame(
  n_locations = nrow(observed_track),
  n_steps = nrow(observed_track) - 1,
  duration_days = as.numeric(diff(range(observed_track$t_)), units = "days")
)
```

```{r observed-track-plot, eval = run_deer_analysis}
ggplot2::ggplot(observed_track, ggplot2::aes(x = x_, y = y_)) +
  ggplot2::geom_path(linewidth = 0.4, color = "firebrick") +
  ggplot2::coord_equal() +
  ggplot2::theme_bw() +
  ggplot2::labs(
    x = "Easting (m)",
    y = "Northing (m)",
    title = "Observed red deer trajectory"
  )
```

## Fit the iSSA with amt

The model specification follows the empirical analysis in the article:

```{r prepare-steps, eval = run_deer_analysis}
ssf_data <- observed_track |>
  amt::steps() |>
  amt::random_steps(n_control = 25) |>
  amt::extract_covariates(forest_map)

ssf_data$cos_ta <- cos(ssf_data$ta_)
ssf_data$log_sl <- log(ssf_data$sl_)
ssf_data <- stats::na.omit(ssf_data)

table(ssf_data$case_)
```

```{r sampling-design, eval = run_deer_analysis}
used_steps <- ssf_data[ssf_data$case_, ]
available_steps <- ssf_data[!ssf_data$case_, ]

ggplot2::ggplot() +
  ggplot2::geom_segment(
    data = available_steps,
    ggplot2::aes(x = x1_, y = y1_, xend = x2_, yend = y2_),
    color = "grey75",
    alpha = 0.25,
    linewidth = 0.2
  ) +
  ggplot2::geom_segment(
    data = used_steps,
    ggplot2::aes(x = x1_, y = y1_, xend = x2_, yend = y2_),
    color = "firebrick",
    linewidth = 0.5
  ) +
  ggplot2::coord_equal() +
  ggplot2::theme_bw() +
  ggplot2::labs(
    x = "Easting (m)",
    y = "Northing (m)",
    title = "Used and available steps",
    subtitle = "Observed steps in red, available control steps in grey"
  )
```

```{r fit-issf, eval = run_deer_analysis}
deer_issf <- amt::fit_issf(
  ssf_data,
  case_ ~ forest + sl_ + log_sl + cos_ta + strata(step_id_)
)

summary(deer_issf)
```

## Simulate from the fitted model

`gmov` validates simulated tracks supplied by the user. The helper below is
included only to make this vignette self-contained. It mirrors the simulation
logic used in the article: draw movement candidates from fitted movement
distributions, evaluate the fitted iSSA selection weights, and sample one next
location at each time step. The publication-scale analysis uses 99 simulations;
the environment variable `GMOV_DEER_N_SIMS` controls the number used when this
vignette is rendered.

```{r simulation-helper, eval = run_deer_analysis}
coef_or_zero <- function(beta, candidates) {
  hit <- candidates[candidates %in% names(beta)]
  if (length(hit) < 1L) {
    return(0)
  }
  unname(beta[hit[1]])
}

simulate_deer_issf <- function(
  model,
  observed_track,
  map,
  env_covar_name = "forest",
  n_sims = 9,
  n_cands = 25
) {
  raw_model <- if (inherits(model, "fit_clogit")) model$model else model
  beta_mean <- stats::coef(raw_model)
  beta_vcov <- stats::vcov(raw_model)
  beta_draws <- mvtnorm::rmvnorm(
    n = n_sims,
    mean = beta_mean,
    sigma = beta_vcov
  )
  colnames(beta_draws) <- names(beta_mean)

  observed_steps <- stats::na.omit(amt::steps(observed_track))
  sl_distr <- amt::fit_distr(observed_steps$sl_, "gamma")
  ta_distr <- amt::fit_distr(observed_steps$ta_, "vonmises")

  n_steps <- nrow(observed_track) - 1L
  x0 <- observed_track$x_[1]
  y0 <- observed_track$y_[1]
  angle0 <- atan2(
    observed_track$y_[2] - observed_track$y_[1],
    observed_track$x_[2] - observed_track$x_[1]
  )

  lapply(seq_len(n_sims), function(i) {
    beta_i <- beta_draws[i, ]
    beta_env <- coef_or_zero(beta_i, env_covar_name)
    beta_sl <- coef_or_zero(beta_i, "sl_")
    beta_log_sl <- coef_or_zero(beta_i, c("log_sl", "log_sl_"))
    beta_cos_ta <- coef_or_zero(beta_i, c("cos_ta", "cos_ta_"))

    x <- numeric(n_steps + 1L)
    y <- numeric(n_steps + 1L)
    x[1] <- x0
    y[1] <- y0
    current_angle <- angle0
    last_step <- n_steps

    for (step_i in seq_len(n_steps)) {
      candidate_sl <- amt::random_numbers(sl_distr, n = n_cands)
      candidate_ta <- amt::random_numbers(ta_distr, n = n_cands)
      candidate_angle <- current_angle + candidate_ta
      candidate_x <- x[step_i] + candidate_sl * cos(candidate_angle)
      candidate_y <- y[step_i] + candidate_sl * sin(candidate_angle)

      env_values <- terra::extract(map, cbind(candidate_x, candidate_y))[[env_covar_name]]
      valid <- is.finite(env_values) & candidate_sl > 0

      if (!any(valid)) {
        last_step <- step_i - 1L
        break
      }

      log_weight <- beta_env * env_values +
        beta_sl * candidate_sl +
        beta_log_sl * log(candidate_sl) +
        beta_cos_ta * cos(candidate_ta)
      log_weight[!valid] <- -Inf

      weight <- exp(log_weight - max(log_weight, na.rm = TRUE))
      weight[!is.finite(weight)] <- 0

      if (sum(weight) <= 0) {
        choice <- sample(which(valid), size = 1L)
      } else {
        choice <- sample.int(n_cands, size = 1L, prob = weight)
      }

      x[step_i + 1L] <- candidate_x[choice]
      y[step_i + 1L] <- candidate_y[choice]
      current_angle <- candidate_angle[choice]
    }

    data.frame(
      x_ = x[seq_len(last_step + 1L)],
      y_ = y[seq_len(last_step + 1L)]
    )
  })
}
```

```{r simulate-tracks, eval = run_deer_analysis}
simulated_tracks <- simulate_deer_issf(
  model = deer_issf,
  observed_track = observed_track,
  map = forest_map,
  env_covar_name = "forest",
  n_sims = deer_n_sims,
  n_cands = 25
)

length(simulated_tracks)
```

```{r simulated-tracks-plot, eval = run_deer_analysis}
simulated_df <- do.call(
  rbind,
  Map(
    function(track, id) {
      data.frame(sim_id = id, x_ = track$x_, y_ = track$y_)
    },
    simulated_tracks,
    seq_along(simulated_tracks)
  )
)

ggplot2::ggplot() +
  ggplot2::geom_path(
    data = simulated_df,
    ggplot2::aes(x = x_, y = y_, group = sim_id),
    color = "grey65",
    alpha = 0.45,
    linewidth = 0.25
  ) +
  ggplot2::geom_path(
    data = observed_track,
    ggplot2::aes(x = x_, y = y_),
    color = "firebrick",
    linewidth = 0.6
  ) +
  ggplot2::coord_equal() +
  ggplot2::theme_bw() +
  ggplot2::labs(
    x = "Easting (m)",
    y = "Northing (m)",
    title = "Observed and simulated red deer trajectories",
    subtitle = "Observed trajectory in red, simulations in grey"
  )
```

## Validate the simulated tracks with gmov

For the red deer empirical application, we validate the first three pillars.
There is no barrier diagnostic because the article did not specify a known
barrier geometry for this individual.

```{r validate-deer, eval = run_deer_analysis}
deer_validation <- validate_ssf_generative(
  observed = observed_track,
  simulated = simulated_tracks,
  metrics = c("ud", "msd", "sinuosity"),
  ud_args = list(grid_size = 30),
  msd_args = list(max_lag = min(120, nrow(observed_track) - 1L))
)

summary(deer_validation)
```

```{r plot-deer-ud, eval = run_deer_analysis}
plot(deer_validation, metric = "ud")
```

```{r plot-deer-msd, eval = run_deer_analysis}
plot(deer_validation, metric = "msd")
```

```{r plot-deer-sinuosity, eval = run_deer_analysis}
plot(deer_validation, metric = "sinuosity")
```

## Reading the diagnostics

The interpretation follows the article. Passing one pillar does not imply that
the fitted iSSA is generatively realistic in every respect. For the red deer
case study, the article reports that the fitted model reproduced broad space
use and path straightness, but showed a detectable mismatch in displacement
dynamics. Ecologically, that pattern suggests that a single-state iSSA may
average over heterogeneous movement phases: it can place simulated trajectories
in broadly the right part of the landscape while still getting the temporal
organization of movement wrong.

The exact numerical values produced by this vignette may differ from the
publication tables because the number of simulations is configurable and the
current package implementation uses the diagnostics documented in `gmov`.
Publication-scale analyses should use a fixed random seed, `K = 99` or larger,
and the same simulation design as the analysis being reported.

## References

Avgar, T., Potts, J. R., Lewis, M. A., and Boyce, M. S. (2016). Integrated
step selection analysis: bridging the gap between resource selection and animal
movement. *Methods in Ecology and Evolution*, 7, 619-630.
<https://doi.org/10.1111/2041-210X.12528>

Nicosia, A. (2026). Beyond the next step: A multi-criteria generative validation
framework for step selection functions. *Methods in Ecology and Evolution*.
<https://doi.org/10.1111/2041-210x.70313>

Signer, J., Fieberg, J., and Avgar, T. (2019). Animal movement tools (`amt`):
R package for managing tracking data and conducting habitat selection analyses.
*Ecology and Evolution*, 9, 880-890. <https://doi.org/10.1002/ece3.4823>
