---
title: "Generative validation of SSF / iSSF models"
output: rmarkdown::html_vignette
embed-resources: true
vignette: >
  %\VignetteIndexEntry{Generative validation of SSF / iSSF models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4,
  message = FALSE,
  warning = FALSE
)
```

```{r setup}
library(gmov)
```

## Motivation

Step-selection functions (SSFs) and integrated step-selection analyses (iSSAs)
are usually fitted to local, one-step movement decisions. A model can therefore
look satisfactory at the scale at which it was fitted while still generating
unrealistic trajectories when it is iterated forward. Generative validation
asks a direct question: if we simulate full trajectories from the fitted model,
do the simulated tracks reproduce important trajectory-level patterns in the
observed track?

Nicosia (2026) proposes a multi-criteria framework with four complementary
pillars:

1. Emergent space use, comparing observed and simulated utilization
   distributions.
2. Diffusion behavior, comparing mean squared displacement (MSD) curves.
3. Path structure, comparing path straightness or sinuosity.
4. Barrier interactions, comparing intersections with a known linear barrier.

`gmov` implements these diagnostics for observed and simulated tracks. It is
designed to sit after a modelling workflow such as `amt`: users fit an SSF or
iSSF, simulate replicate tracks, and then pass those tracks to `gmov`.

## Red deer empirical example

The empirical application in Nicosia (2026) uses the red deer trajectory and
forest raster distributed with `amt` (Signer, Fieberg, and Avgar, 2019). The
article fits a standard iSSA to a GPS-tracked red deer from northern Germany,
using a 6-hour sampling interval, a binary forest covariate, and a 1:25
used-to-available step design. The fitted model includes forest selection and
movement-kernel terms:

```r
case_ ~ forest + sl_ + log_sl + cos_ta + strata(step_id_)
```

Because the empirical example does not specify a known barrier geometry, the
article evaluates the first three pillars: emergent space use, MSD, and
sinuosity.

This vignette ships a compact, precomputed version of that red deer validation
example so that the vignette renders with visible results on installation and
on GitHub. The object contains the observed red deer track, 99 simulated tracks
from the empirical iSSA workflow used in the article, and the scorecard reported
by that workflow. The full `amt` fitting code is shown later for context.

## Load the validation ensemble

```{r load-data}
data("red_deer_gmov", package = "gmov")

observed_track <- red_deer_gmov$observed_track
simulated_tracks <- red_deer_gmov$simulated_tracks

data.frame(
  observed_locations = nrow(observed_track),
  observed_steps = nrow(observed_track) - 1,
  simulated_tracks = length(simulated_tracks),
  simulated_locations_min = min(vapply(simulated_tracks, nrow, integer(1))),
  simulated_locations_max = max(vapply(simulated_tracks, nrow, integer(1)))
)
```

The table below is the scorecard from the empirical workflow in Nicosia (2026).
It is included as provenance for the precomputed simulation ensemble.

```{r article-scorecard}
red_deer_gmov$article_scorecard
```

The article reports a pattern of partial generative realism: the fitted iSSA
reproduces broad space-use and path straightness, but shows a detectable
mismatch in displacement dynamics.

## Observed and simulated trajectories

The observed trajectory is shown in red. A subset of the 99 simulated
trajectories is shown in grey to keep the plot readable.

```{r trajectory-plot}
plot_sims <- simulated_tracks[seq_len(25)]
plot_sims_df <- do.call(
  rbind,
  Map(
    function(track, id) {
      data.frame(sim_id = id, x_ = track$x_, y_ = track$y_)
    },
    plot_sims,
    seq_along(plot_sims)
  )
)

ggplot2::ggplot() +
  ggplot2::geom_path(
    data = plot_sims_df,
    ggplot2::aes(x = x_, y = y_, group = sim_id),
    color = "grey70",
    alpha = 0.45,
    linewidth = 0.25
  ) +
  ggplot2::geom_path(
    data = observed_track,
    ggplot2::aes(x = x_, y = y_),
    color = "firebrick",
    linewidth = 0.7
  ) +
  ggplot2::coord_equal() +
  ggplot2::theme_bw() +
  ggplot2::labs(
    x = "Easting (m)",
    y = "Northing (m)",
    title = "Observed and simulated red deer trajectories",
    subtitle = "Observed track in red; 25 simulated tracks in grey"
  )
```

## Run the gmov diagnostics

The call below validates the same supplied simulation ensemble with the current
`gmov` diagnostics. The UD diagnostic uses an empirical grid approximation to
the utilization distribution. The MSD diagnostic uses lagged MSD curves. These
implementation details are documented in the package help pages, and numerical
values can differ from the prototype diagnostics used in the article.

```{r run-validation}
deer_validation <- validate_ssf_generative(
  observed = observed_track,
  simulated = simulated_tracks,
  metrics = c("ud", "msd", "sinuosity"),
  ud_args = list(grid_size = 20),
  msd_args = list(max_lag = 80)
)

summary(deer_validation)
```

The result is an object of class `gmov_generative`.

```{r print-validation}
deer_validation
```

## Visual diagnostics

The UD diagnostic compares the observed grid utilization distribution to the
simulated grid utilization distributions using the 1-Wasserstein distance
between empirical grid masses.

```{r plot-ud}
plot(deer_validation, metric = "ud")
```

The MSD diagnostic compares the observed lagged displacement curve to the
simulation envelope. This targets temporal organization of movement, not only
the final spatial footprint.

```{r plot-msd}
plot(deer_validation, metric = "msd")
```

The sinuosity diagnostic compares the observed straightness index with the
distribution of straightness indices from simulated trajectories.

```{r plot-sinuosity}
plot(deer_validation, metric = "sinuosity")
```

## Reproducing the model-fitting workflow with amt

The following code shows the `amt` workflow corresponding to the empirical
application. It is not evaluated in this vignette because the package vignette
uses the precomputed simulation ensemble above. Running the full workflow
requires a working `amt` installation and can take longer than a typical package
vignette build.

```{r amt-fit-workflow, eval = FALSE}
library(amt)
library(terra)
library(survival)

set.seed(2025)

data("deer", package = "amt")
data("sh_forest", package = "amt")

forest_map <- terra::rast(sh_forest)
names(forest_map) <- "forest"

observed_track <- deer |>
  track_resample(rate = hours(6), tolerance = minutes(30)) |>
  filter_min_n_burst(min_n = 2)

observed_track <- observed_track[, c("x_", "y_", "t_")]

ssf_data <- observed_track |>
  steps() |>
  random_steps(n_control = 25) |>
  extract_covariates(forest_map)

ssf_data$cos_ta <- cos(ssf_data$ta_)
ssf_data$log_sl <- log(ssf_data$sl_)
ssf_data <- stats::na.omit(ssf_data)

deer_issf <- fit_issf(
  ssf_data,
  case_ ~ forest + sl_ + log_sl + cos_ta + strata(step_id_)
)

summary(deer_issf)
```

After fitting, users should simulate replicate tracks from the fitted model
using the simulation procedure appropriate for their analysis. Once the
simulation output is a list of track-like objects, the validation step is the
same as above:

```{r gmov-after-amt, eval = FALSE}
deer_validation <- validate_ssf_generative(
  observed = observed_track,
  simulated = simulated_tracks,
  metrics = c("ud", "msd", "sinuosity"),
  ud_args = list(grid_size = 20),
  msd_args = list(max_lag = 80)
)

summary(deer_validation)
plot(deer_validation, metric = "msd")
```

## Interpretation

The red deer example illustrates why generative validation is useful even when
a fitted iSSA is plausible at the one-step scale. The model can generate
trajectories that occupy broadly similar regions of space while still failing
to reproduce the temporal structure of displacement. In the article workflow,
this appears as a mismatch in the MSD pillar, while the broad space-use and
sinuosity pillars are not rejected.

In practice, the diagnostics should be read together. A single p-value is not a
complete model assessment. Instead, the pattern across pillars helps identify
which aspect of generative realism is plausible and which aspect may require a
better movement kernel, additional behavioural states, temporal structure, or
landscape constraints.

## References

Avgar, T., Potts, J. R., Lewis, M. A., and Boyce, M. S. (2016). Integrated
step selection analysis: bridging the gap between resource selection and animal
movement. *Methods in Ecology and Evolution*, 7, 619-630.
<https://doi.org/10.1111/2041-210X.12528>

Nicosia, A. (2026). Beyond the next step: A multi-criteria generative validation
framework for step selection functions. *Methods in Ecology and Evolution*.
<https://doi.org/10.1111/2041-210x.70313>

Signer, J., Fieberg, J., and Avgar, T. (2019). Animal movement tools (`amt`):
R package for managing tracking data and conducting habitat selection analyses.
*Ecology and Evolution*, 9, 880-890. <https://doi.org/10.1002/ece3.4823>
