---
title: "Generative validation of SSF / iSSF models"
output: rmarkdown::html_vignette
embed-resources: true
vignette: >
  %\VignetteIndexEntry{Generative validation of SSF / iSSF models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
```

```{r setup}
library(gmov)
```

## Motivation

SSF and iSSF models are usually fitted to one-step transitions. A model can fit
local movement decisions while still generating unrealistic long-run
trajectories when iterated forward. The goal of `gmov` is to validate simulated
trajectories against observed trajectories using trajectory-level summaries.

The framework follows Nicosia (2026), which proposes four complementary
generative validation pillars:

- emergent space use;
- diffusion behavior;
- path structure;
- barrier interactions.

This vignette uses simulated toy data. It is illustrative and does not
reproduce the empirical or synthetic results from the article.

All examples below use planar toy coordinates. For real data, observed tracks,
simulated tracks, and barrier geometries should use a common projected
coordinate system.

Reference: Nicosia, A. (2026). Beyond the next step: A multi-criteria
generative validation framework for step selection functions. *Methods in
Ecology and Evolution*. <https://doi.org/10.1111/2041-210x.70313>

## Toy tracks

```{r toy-data}
set.seed(10)

simulate_track <- function(n = 60, step_sd = 1, drift = c(0, 0)) {
  dx <- stats::rnorm(n, mean = drift[1], sd = step_sd)
  dy <- stats::rnorm(n, mean = drift[2], sd = step_sd)

  data.frame(
    x = cumsum(dx),
    y = cumsum(dy)
  )
}

observed_track <- simulate_track(n = 60, step_sd = 0.8)
simulated_tracks <- replicate(
  n = 9,
  expr = simulate_track(n = 60, step_sd = 0.8),
  simplify = FALSE
)
```

## Run the validation

```{r validation}
res <- validate_ssf_generative(
  observed = observed_track,
  simulated = simulated_tracks,
  metrics = c("ud", "msd", "sinuosity"),
  ud_args = list(grid_size = 12),
  msd_args = list(max_lag = 20)
)

summary(res)
```

## Visual summaries

```{r msd-plot}
plot(res, metric = "msd")
```

```{r sinuosity-plot}
plot(res, metric = "sinuosity")
```

## Barrier interactions

Barrier validation requires a known linear feature supplied before validation.
The example below creates a simple vertical line. The crossing diagnostic tests
whether observed crossings are unusually few relative to simulations by
default, which corresponds to a fitted model that may be too permeable.

```{r barrier}
barrier <- sf::st_sfc(
  sf::st_linestring(matrix(c(0, -20, 0, 20), ncol = 2, byrow = TRUE)),
  crs = 3857
)

barrier_res <- validate_ssf_generative(
  observed = observed_track,
  simulated = simulated_tracks,
  metrics = "barrier",
  barrier = barrier
)

summary(barrier_res)
plot(barrier_res, metric = "barrier")
```

## Sketch of an amt workflow

The package is intended to sit after an `amt` modeling and simulation workflow.
The precise simulation call depends on the fitted model and user choices, so
the outline below is not evaluated.

```{r amt-sketch, eval = FALSE}
library(amt)
library(gmov)

# observed_track <- ...
# steps <- observed_track |>
#   steps_by_burst() |>
#   random_steps(n = 15)
#
# fitted_model <- steps |>
#   fit_clogit(case_ ~ forest + log(sl_) + cos(ta_) + strata(step_id_))
#
# simulated_tracks <- ...

res <- validate_ssf_generative(
  observed = observed_track,
  simulated = simulated_tracks,
  metrics = c("ud", "msd", "sinuosity")
)
```
